---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Data are loaded using *read.csv*. Dates are kept as dates (and not changed to factors).

```{r}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
activity$date <- as.POSIXct(activity$date)
str(activity)
```



## What is mean total number of steps taken per day?

Function *aggregate* is used sum up steps per day.

```{r}
activity_day <- aggregate(activity["steps"], by = activity["date"], FUN = sum)
max_steps <- max(activity_day$steps, na.rm = TRUE)
min_steps <- min(activity_day$steps, na.rm = TRUE)
mean_steps <- format(mean(activity_day$steps, na.rm = TRUE), digits = 7, scientific = FALSE)
med_steps <- median(activity_day$steps, na.rm = TRUE)
```

The figure below shows the distribution of the total number of steps taken per day.

```{r}
hist(activity_day[["steps"]], breaks = 10, 
     main = "Distribution of Total Number of Steps per Day",
     xlab = "Steps per Day")
```

The minimum number of steps per day is `r min_steps` and the maximum number  is `r max_steps`. The mean number is `r mean_steps` and the median is `r med_steps`. (All numbers calculated while ignoring NAs). 


## What is the average daily activity pattern?

Function *aggregate* is used to calculate the mean number of steps per interval.

```{r}
activity_int <- aggregate(activity["steps"], by = activity["interval"], FUN = mean, na.rm = TRUE)
int_max_steps <- activity_int[match(max(activity_int$steps), activity_int$steps),1]
```

The figure below shows the daily activity pattern as mean number of steps per interval. Labels on the x-axis show strting time of intervals in format *hhmm*.

```{r}
plot(activity_int, type = "l",
    main = "Mean Number of Steps per Five Minute Time Interval",
    xlab = "Time", ylab =  "Steps")
```

The maximum number of steps for an interval is seen for interval `r int_max_steps`.


## Imputing missing values

The number of missing values in the data set may easily be read from *str(activity)*. Here an alternativa approach is shown.

```{r}
missingValues <- nrow(activity) - nrow(na.omit(activity))
missingSteps <- sum(is.na(activity$steps))
missingDate <- sum(is.na(activity$date))
missingInt <- sum(is.na(activity$interval))
```

The number of missing values (rows with NA) is `r missingValues`. All missing values relate to column **steps**, as missing values for **date** and **interval** are `r missingDate` and `r missingInt` respectively. 

### Imputing strategy

Missing values are replaced by the mean number of steps for day of week and interval.

### Imputation

First build build dataset *activityWD_int* containing means of steps for day of week and interval.

```{r}
activityWD <- activity
activityWD$wkday <- weekdays(activityWD$date, abbreviate = TRUE)
activityWD_int <- aggregate(activityWD["steps"], 
                            by = activityWD[c("wkday","interval")], 
                            FUN = mean, na.rm = TRUE)
activityWD_int$wkday_int <- paste0(activityWD_int$wkday, "_", activityWD_int$interval)
```

Next build dataset *activityNoNA* with imputed values.

```{r}
activityNoNA <- activity
for (i in seq_along(activityNoNA$steps)) {
    if (is.na(activityNoNA[i,"steps"])) {
        activityNoNA[i, "steps"] <- 
            activityWD_int[match(paste0(weekdays(activityNoNA[i, "date"], abbreviate = TRUE),"_", 
                                        activityNoNA[i, "interval"]), 
                                 activityWD_int$wkday_int), "steps"]
    }
}
```


Function *aggregate* is used sum up steps per day.

```{r}
activity_day <- aggregate(activityNoNA["steps"], by = activityNoNA["date"], FUN = sum)
max_steps_NoNA <- format(max(activity_day$steps, na.rm = TRUE), digits = 7, scientific = FALSE)
min_steps_NoNA <- min(activity_day$steps, na.rm = TRUE)
mean_steps_NoNA <- format(mean(activity_day$steps, na.rm = TRUE), digits = 7, scientific = FALSE)
med_steps_NoNA <- format(median(activity_day$steps, na.rm = TRUE), digits = 7, scientific = FALSE)
```

The figure below shows the distribution of the total number of steps taken per day.

```{r}
hist(activity_day[["steps"]], breaks = 10, 
     main = "Distribution of Total Number of Steps per Day",
     xlab = "Steps per Day")
```

The minimum number of steps per day is `r min_steps_NoNA` and the maximum number  is `r max_steps_NoNA`. The mean number is `r mean_steps_NoNA` and the median is `r med_steps_NoNA`. 

## Are there differences in activity patterns between weekdays and weekends?
